---
title: How to load a gene model from a GFF or GTF file
author: Bioconductor Core Team
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to load a gene model from a GFF or GTF file}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk: 
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

A **gene model** is essentially a set of annotations that
describes the genomic locations of the known genes, transcripts, exons, and CDS, for
a given organism. The a standardized file format to hold gene models is a **[GFF or GTF](https://useast.ensembl.org/info/website/upload/gff.html)**. In Bioconductor, gene model information is typically represented as a TxDb object but also sometimes as a GRanges or GRangesList object. We can use the `makeTxDbFromGFF()` function from the *[txdbmaker](https://bioconductor.org/packages/release/bioc/html/txdbmaker.html)* package to import a GFF or GTF file as a *TxDb* object.


# Bioconductor packages used in this document

* `r Biocpkg("txdbmaker")`

# How to load a gene model from a GFF or GTF file

We will use a small .gff3 file provided by the *[txdbmaker](https://bioconductor.org/packages/release/bioc/html/txdbmaker.html)* package.

```{r, warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
    library(txdbmaker)
})

 gff_file <- system.file("extdata", "GFF3_files", "a.gff3", package="txdbmaker")
 
 txdb <- makeTxDbFromGFF(gff_file, format="gff3")
 txdb

```


See `?makeTxDbFromGFF` in the *[txdbmaker](https://bioconductor.org/packages/release/bioc/html/txdbmaker.html)* package for more information.

Extract the exon coordinates grouped by gene from this gene model:

```{r}

 exonsBy(txdb, by="gene")


```


# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>

