---
title: How to use tidy principles for genomic ranges manipulation
author: Jacques Serizay
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to use tidy principles for genomic ranges manipulation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk: 
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
bibliography: ../inst/references.bib
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

Tidy data principles are a set of principles for structuring data in a way 
that makes it easier to manipulate and analyze. They have been popularized by the
`tidyverse` collection of R packages, which includes `dplyr` for data manipulation 
and `ggplot2` for data visualization. The tidy principles can also be applied to
biological data, such as RNA-seq data, to facilitate analysis and visualization, 
thanks to the `tidyomics` meta-package.

# Bioconductor packages used in this document

* `r Biocpkg("tidyomics")`

# How to manipulate GRanges objects with tidy principles

For illustration, we are going to download a `xlsx` file provided as a 
Supplementary Data in [@Serizay2020Oct], and manipulate it using tidy principles.

Good practices recommended by the Bioconductor consortium suggest to use 
the `BiocFileCache` package to download and cache files from remote servers.

```{r}
library(BiocFileCache)

# Create a BiocFileCache instance
bfc <- BiocFileCache()

# Download the xlsx file from Ensembl and cache it locally using BiocFileCache
xlsx_url <- "https://genome.cshlp.org/content/suppl/2020/11/16/gr.265934.120.DC1/Supplemental_Table_S2.xlsx"
xlsx_file <- bfcrpath(bfc, xlsx_url)

# The file is now downloaded and available
xlsx_file
```

Excel files can be read into `tibbles` in `R` thanks to the `read_xlsx()` function 
from the `readxl` package.

```{r}
library(readxl)
tbl <- read_xlsx(xlsx_file, sheet = "ATAC_metrics", skip = 2)
tbl
```

We convert it into a `GRanges` object directly using `as_granges()` from the `plyranges` package, 
using [tidy evaluation as in `dplyr` package](https://dplyr.tidyverse.org/articles/programming.html). 

```{r}
library(tidyomics)
gr <- tbl |> 
    as_granges(seqnames = chrom_ce11, start = start_ce11, end = end_ce11)
gr
```

Let's say we are not interested in all the metadata columns, and we want to keep only the
`annot_dev_age_tissues`, `Annotation`, and the columns ending with `_RPM`. 
We can use the `select()` function from `dplyr` to select these columns, 
just like we would do with a `data.frame` or `tibble`.

```{r}
gr <- gr |> 
    select(annot_dev_age_tissues, Annotation, ends_with("_RPM"))
gr
```

In fact, most verbs from the `dplyr` package can be used with `GRanges` objects,
such as `filter()`, `mutate()`, `arrange()`, `group_by()` and `summarise()`. 

```{r}
gr |> 
    mutate(start = start - 1000, end = end + 1000) |>
    filter(start > 1e6) |>
    arrange(Intest._RPM) |>
    group_by(annot_dev_age_tissues, Annotation) |> 
    summarize(n = plyranges::n()) 
```

Note that the `summarize()` function automatically returns a `DataFrame`, not a `GRanges` object. 

To enhance compatibility with the `tidyverse` ecosystem, the `plyranges` package
provides a `as_tibble()` method for `GRanges` objects, which converts them
into `tibbles`. This allows you to use pass `GRanges` objects to `ggplot` for visualization. 

```{r}
gr |> 
    filter(Annotation %in% c("Intest.", "Hypod.", "Soma", "Ubiq.")) |>
    as_tibble() |> 
    ggplot(aes(x = `Intest._RPM`, y = `Hypod._RPM`, color = Annotation)) +
    geom_point() +
    labs(x = "Intestine RPM", y = "Hypodermis RPM") + 
    facet_grid(Annotation~annot_dev_age_tissues) +
    theme_bw() +
    theme(legend.position = "bottom")
```

Finally, the `join` concept introduced by the `dplyr` package can also be applied to `GRanges` objects.
In this case, the operation refers to joining overlapping, nearest, or preceding/following ranges, 
using the `join_*_*()` function from the `plyranges` package.

As an example, we can compare annotations from [@Serizay2020Oct] to chromosome 
arms/center 

```{r}
arms <- GRanges(
    seqnames = rep(c("chrI", "chrII", "chrIII", "chrIV", "chrV", "chrX"), each = 3),
    ranges = IRanges(
        start = c(
            1, 4550001, 10750001, 
            1, 4550001, 10750001, 
            1, 4450001, 10150001, 
            1, 4750001, 12650001, 
            1, 4850001, 16250001, 
            1, 2850001, 14250001
        ), 
        end = c(
            4550000, 10750000, 15072423, 
            4550000, 10750000, 15279345,
            4450000, 10150000, 13783700,
            4750000, 12650000, 17493793,
            4850000, 16250000, 20924149,
            2850000, 14250000, 17718866
        )
    ),
    strand = "*",
    part = rep(c("arm", "center", "arm"), 6)
)

gr |> 
    join_overlap_left(arms) |> 
    group_by(Annotation, part) |>
    filter(Annotation %in% c("Germline", "Neurons")) |>
    summarize(n = plyranges::n()) 
```

# Further reading 

To read more about the `tidyomics` packages, please refer to the
[tidyomics](https://github.com/tidyomics) GitHub organization page. 

More tutorials are provided by the `tidyomics` GitHub organization in the 
[tidy-ranges-tutorial](https://tidyomics.github.io/tidy-ranges-tutorial/) website. 

# References

::: {#refs}
:::

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>

