---
title: How to use tidy principles for RNA-seq data analysis
author: Bioconductor Core Team
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to use tidy principles for RNA-seq data analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk: 
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

# Bioconductor packages used in this document

* `r Biocpkg("tidyomics")`

# How to manipulate SummarizedExperiment objects with tidy principles

For illustration, we use data from the `r Biocpkg("tidySummarizedExperiment")` 
package (included in the meta-package `r Biocpkg("tidyomics")`). 

```{r}
library(tidyomics)

pasilla_tidy <- tidySummarizedExperiment::pasilla 
pasilla_tidy
```

How `pasilla_tidy` object is printed is controlled by the `tidySummarizedExperiment` package, 
and by default it is displayed as a `tibble`-like object. Note that this 
behavior can be changed by setting the `options(restore_SummarizedExperiment_show = TRUE)` option. 

```{r}
# Turn off the tibble visualisation
options("restore_SummarizedExperiment_show" = TRUE)
pasilla_tidy

# Turn on the tibble visualisation
options("restore_SummarizedExperiment_show" = FALSE)
pasilla_tidy
```

The `pasilla_tidy` object is a `SummarizedExperiment` object, and the 
`tidySummarizedExperiment` packages adds an invisible layer to abstract the `SE` object as a
`tibble`. This allows us to use the `dplyr` package to manipulate the data, and eventually 
to directly plug data flow into `ggplot2` for visualization.

```{r}
pasilla_tidy |> 
    filter(condition == "untreated", type == "single_end") |> 
    select(-type) |> 
    group_by(.feature, .sample) |> 
    summarise(tot_counts = sum(counts)) |>
    ggplot(aes(x = .sample, y = tot_counts+1, fill = .sample)) + 
    geom_violin() + 
    geom_boxplot(outliers = FALSE, width = 0.05, fill = "white") +
    scale_y_log10() + 
    annotation_logticks(side = 'l') +
    labs(x = "Sample", y = "Total counts (log1p)") +
    theme_bw()
```

Because these tidy methods are defined for `SummarizedExperiment` objects, 
most classes built on top of `SummarizedExperiment`, such as `DESeqDataSet` or `SingleCellExperiment`, 
can also be manipulated with tidy principles. This has prompted the development of 
other packages (included in the `tidyomics` meta-package) that extend the tidy principles to
other analyses, such as single-cell transcriptomics analysis.

```{r}
data(pbmc_small, package="tidySingleCellExperiment")
pbmc_small

## We can still manipulate the data with standard SummarizedExperiment methods
counts(pbmc_small)[1:5, 1:4]

## But we can also use tidy principles
pbmc_small |> 
    extract(file, "sample", "../data/([a-z0-9]+)/outs.+") |> 
    rename(annotation = letter.idents) |> 
    ggplot(aes(sample, nCount_RNA, fill=sample)) +
    geom_boxplot(outlier.shape=NA) +
    geom_jitter(width=0.1) +
    facet_wrap( ~ annotation)
```

To read more about the `tidyomics` packages, please refer to the
[tidyomics](https://github.com/tidyomics) GitHub organization page. 

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>

