---
title: How to retrieve a gene model from AnnotationHub
author: Bioconductor Core Team
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to retrieve a gene model from AnnotationHub}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

This HowTo has been adapted from the list of HowTos provided in the 
[vignette](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesHOWTOs.pdf)
for the `r Biocpkg("GenomicRanges")` Bioconductor package. 

# Bioconductor packages used in this document

* `r Biocpkg("AnnotationHub")`

# How to retrieve a gene model from *AnnotationHub*

When a gene model is not available as a `GRanges` or `GRangesList` object or as 
a \Bioconductor{} data package, it may be available on `r Biocpkg("AnnotationHub")`.
In this *HOWTO*, will look for a gene model for Drosophila melanogaster
on `r Biocpkg("AnnotationHub")`. Create a `hub' and then filter on Drosophila 
melanogaster:

```{r}
suppressPackageStartupMessages({
    library(AnnotationHub)
})
### Internet connection required!
hub <- AnnotationHub()
hub <- subset(hub, hub$species=='Drosophila melanogaster')
```

There are `r length(hub)` files that match Drosophila melanogaster. 
If you look at the metadata in hub, you can see that the 5th record 
represents a GRanges object from UCSC RefSeq Gene track.

```{r}
length(hub)
head(names(hub))
head(hub$title, n=10)
## then look at a specific slice of the hub object.
hub[5]
```
So you can retrieve that dm3 file as a GRanges like this:

```{r}
gr <- hub[[names(hub)[5]]]
summary(gr)
```

The metadata fields contain the details of file origin and content.

```{r}
metadata(gr)
```
Split the `GRanges` object by gene name to get a `GRangesList` object of 
transcript ranges grouped by gene.

```{r}
txbygn <- split(gr, gr$name)
```

You can now use `txbygn` with the `summarizeOverlaps` function
to prepare a table of read counts for RNA-Seq differential gene expression.

# Further reading

Note that before passing `txbygn` to `summarizeOverlaps`,
you should confirm that the seqlevels (chromosome names) in it match those
in the BAM file. See `?renameSeqlevels`, `?keepSeqlevels`
and `?seqlevels` for examples of renaming seqlevels.

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>
