---
title: How to use tidy principles for RNA-seq data analysis
author: Jacques Serizay
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to use tidy principles for RNA-seq data analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk: 
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

Tidy data principles are a set of principles for structuring data in a way 
that makes it easier to manipulate and analyze. They have been popularized by the
`tidyverse` collection of R packages, which includes `dplyr` for data manipulation 
and `ggplot2` for data visualization. The tidy principles can also be applied to
biological data, such as RNA-seq data, to facilitate analysis and visualization, 
thanks to the `tidyomics` meta-package.

# Bioconductor packages used in this document

* `r Biocpkg("tidyomics")`

# How to manipulate SummarizedExperiment objects with tidy principles

For illustration, we use data from the `r Biocpkg("airway")` package. 

```{r}
library(tidyomics)
data(airway, package="airway")

airway
```

How `airway` object is printed is controlled by the `tidySummarizedExperiment` package, 
and by default it is displayed as a `tibble`-like object. Note that this 
behavior can be changed by setting the `options(restore_SummarizedExperiment_show = TRUE)` option. 

```{r}
# Turn off the tibble visualisation
options("restore_SummarizedExperiment_show" = TRUE)
airway

# Turn on the tibble visualisation
options("restore_SummarizedExperiment_show" = FALSE)
airway
```

The `airway` object is a `SummarizedExperiment` object, and the 
`tidySummarizedExperiment` packages adds an invisible layer to abstract the `SE` object as a
`tibble`. This allows us to use the `dplyr` package to manipulate the data, and eventually 
to directly plug data flow into `ggplot2` for visualization.

```{r}
airway |> 
    filter(albut == "untrt") |> 
    group_by(.feature, dex) |> 
    summarise(tot_counts = sum(counts)) |>
    pivot_wider(names_from = dex, values_from = tot_counts) |>
    slice_sample(n = 10000) |> 
    ggplot(aes(x = trt, y = untrt)) + 
    geom_density_2d(color = "black", linewidth = 0.5) +
    geom_point(alpha = 0.1, size = 0.3) + 
    scale_x_log10() + 
    scale_y_log10() + 
    annotation_logticks(side = 'bl') +
    labs(x = "Dex treated", y = "Dex untreated") +
    theme_bw()
```

Because these tidy methods are defined for `SummarizedExperiment` objects, 
most classes built on top of `SummarizedExperiment`, such as `DESeqDataSet` or `SingleCellExperiment`, 
can also be manipulated with tidy principles. This has prompted the development of 
other packages (included in the `tidyomics` meta-package) that extend the tidy principles to
other analyses, such as single-cell transcriptomics analysis.

```{r}
data(pbmc_small, package="tidySingleCellExperiment")
pbmc_small

## We can still manipulate the data with standard SummarizedExperiment methods
counts(pbmc_small)[1:5, 1:4]

## But we can also use tidy principles
pbmc_small |> 
    extract(file, "sample", "../data/([a-z0-9]+)/outs.+") |> 
    rename(annotation = letter.idents) |> 
    ggplot(aes(sample, nCount_RNA, fill=sample)) +
    geom_boxplot(outlier.shape=NA) +
    geom_jitter(width=0.1) +
    facet_wrap( ~ annotation)
```

# Further reading 

To read more about the `tidyomics` packages, please refer to the
[tidyomics](https://github.com/tidyomics) GitHub organization page. 

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>

