---
title: How to extract promoter sequences
author: Fabricio Almeida-Silva
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to extract promoter sequences}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
        embed-resources: true
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

Promoters are DNA sequences typically located upstream of a 
gene's transcription start site (TSS), and they contain binding 
sites for transcription factors. 
Researchers working on gene regulation often want to extract
promoter sequences for all genes or a set of genes. With these sequences,
researchers can, for example, look for known transcription factor 
binding sites (TFBS), or predict TFBS *de novo*. This HowTo will demonstrate
how to extract promoter sequences for any gene(s) in a genome.

# Bioconductor packages used in this document

* `r Biocpkg("Biostrings")`
* `r Biocpkg("rtracklayer")`
* `r Biocpkg("GenomicRanges")`
* `r Biocpkg("txdbmaker")`
* `r Biocpkg("GenomicFeatures")`
* `r Biocpkg("BSgenome")`

# How to extract promoter sequences

We will start by loading the required packages.

```{r}
#| message: false

library(Biostrings)
library(GenomicRanges)
library(rtracklayer)
library(txdbmaker)
library(GenomicFeatures)
library(BSgenome)
```

Then, we will get our example data set. Here, we will use data for 
the model plant *Arabidopsis thaliana*. We will start by obtaining 
genome sequences and gene annotation from [Ensembl Plants](https://plants.ensembl.org).

::: {.callout-note}

## Working with model and non-model organisms

Here, we're obtaining data from [Ensembl Plants](https://plants.ensembl.org)
to explicitly show how this can be done for **any** organism. However,
if you're lucky enough to work with a model organism (e.g. human, mouse, etc),
you can find curated genome sequences and gene annotation 
in `r Biocpkg("AnnotationHub")` (see, for instance, 
[this](https://bioconductor.github.io/BiocHowTo/articles/how-to-retrieve-gene-model-from-annotationhub.html) HowTo article).

:::

```{r}
# Read genomic data from Ensembl Plants
## Genome
genome <- Biostrings::readDNAStringSet(
    "https://ftp.ensemblgenomes.ebi.ac.uk/pub/plants/release-61/fasta/arabidopsis_thaliana/dna/Arabidopsis_thaliana.TAIR10.dna.toplevel.fa.gz"
)
names(genome) <- gsub(" .*", "", names(genome)) # remove text after chr name

## Annotation
annotation <- rtracklayer::import(
    "https://ftp.ensemblgenomes.ebi.ac.uk/pub/plants/release-61/gff3/arabidopsis_thaliana/Arabidopsis_thaliana.TAIR10.61.gff3.gz"
)
```

Genome sequences are stored in `DNAStringSet` objects, and 
gene annotation (i.e., start and end coordinates of each gene) are stored in 
`GRanges` objects. This is what these objects look like:

```{r}
# Take a look at the genome and annotation
genome
annotation
```

Note that our `GRanges` object contains ranges for multiple types of features,
including genes, mRNAs, exons, and even entire chromosomes. 
Since we're only interested in promoter regions, we will subset
`annotation` to keep only 'gene' ranges.

```{r}
# Keep only gene ranges
genes <- annotation[annotation$type == "gene"]
genes
```

Next, we can use the `promoters()` function from 
`r Biocpkg("GenomicRanges")` to extract the genomic coordinates of promoters.
By default, promoters start at 2000 bp upstream the transcription start 
site (TSS), and end at 200 bp downstream the TSS, but you can modify these 
positions in parameters *upstream* and *downstream* of `promoters()`. 
For example, some transcription factor families in plants 
(e.g. HSF and C2C2-GATA) bind exclusively to regions downstream the TSS, which
is contrary to the common view of TFBS being located upstream TSSs.
Importantly, if a gene has multiple TSSs (e.g., because of different 
transcripts starting at different positions), the most upstream TSS is 
used, not a cleverly chosen 'representative' or 'canonical' transcript.

```{r}
# Get promoter regions from a `GRanges` object
prom_genes <- GenomicRanges::promoters(genes)

prom_genes
```

Alternatively, if you want to extract promoters for each transcript separately,
you can subset the `GRanges` object to extract only ranges of 
type 'transcript', and then use `promoters()` as demonstrated above.
Note that GFF3 files from some databases do not contain a 'transcript' type.
In these cases, you will most likely find transcript-level ranges in rows of 
type 'mRNA'. This is the case for [Ensembl Plants](https://plants.ensembl.org), 
from where we obtained our example data. 
To extract promoters for each *A. thaliana* transcript, you'd execute:

```{r}
# Extract transcript ranges from a GRanges object
tx_ranges <- annotation[annotation$type == "mRNA"] # or 'transcript' (if any)

# Extract promoters for each transcript separately
prom_tx <- promoters(tx_ranges)
prom_tx
```

::: {.callout-tip}

## Extracting promoter regions from `TxDb` objects

If you have a `TxDb` object with transcript coordinates and metadata,
you can extract promoter regions for each transcript with the `promoters()`
function from the `r Biocpkg("GenomicFeatures")` package. Note that function names
are the same, but `GenomicFeatures::promoters()` takes `TxDb` objects as input, 
while `GenomicRanges::promoters()` takes `GRanges` objects as input.

```{r}
# Create a `TxDb` object from a `GRanges` object
txdb <- txdbmaker::makeTxDbFromGRanges(annotation)

# Get promoter regions for all transcripts
prom_tx2 <- GenomicFeatures::promoters(txdb)

prom_tx2
```

:::

Finally, you can extract sequences given a genome and some coordinates with
the `getSeq()` function from `r Biocpkg("BSgenome")`.

```{r}
# Extract promoter sequences (1k genes only for demo purposes)
prom_seqs <- getSeq(genome, prom_genes[1:1000])

prom_seqs
```


# Further reading

To learn more about how to work with sequences and ranges using Bioconductor 
packages, look at the vignettes of the `r Biocpkg("Biostrings")` and
`r Biocpkg("GenomicRanges")` packages.

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>
