---
title: How to get the exon and intron sequences of a given gene
author: Bioconductor Core Team
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to get the exon and intron sequences of a given gene}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

This HowTo has been adapted from the list of HowTos provided in the 
[vignette](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesHOWTOs.pdf)
for the `r Biocpkg("GenomicRanges")` Bioconductor package. 

# Bioconductor packages used in this document

* `r Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")`
* `r Biocpkg("BSgenome.Hsapiens.UCSC.hg38")`
* `r Biocpkg("GenomicFeatures")`
* `r Biocpkg("Biostrings")`

# How to get the exon and intron sequences of a given gene

The exon and intron sequences of a gene are essentially the DNA sequences of 
the introns and exons of all known transcripts of the gene. The first task, 
therefore, is to identify all transcripts associated with the gene of interest. 
Our example gene is the human `TRAK2`, which is involved in regulation of 
endosome-to-lysosome trafficking of membrane cargo. The Entrez gene id is
'66008'.

```{r}
trak2 <- "66008"
```

The `r Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")` data package contains the 
gene models corresponding to the UCSC 'Known Genes' track.

```{r}
suppressPackageStartupMessages({
    library(TxDb.Hsapiens.UCSC.hg38.knownGene)
})
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
```

The transcript ranges for all the genes in the gene model can be extracted with 
the `transcriptsBy` function from the `r Biocpkg("GenomicFeatures")` package. 
They will be returned in a named `GRangesList` object containing all the 
transcripts grouped by gene. In order to keep only the transcripts of the 
`TRAK2` gene we will subset the `GRangesList` object using the `[[` operator.

```{r}
suppressPackageStartupMessages({
    library(GenomicFeatures)
})
(trak2_txs <- transcriptsBy(txdb, by = "gene")[[trak2]])
```

`trak2_txs` is a `GRanges` object with one range per transcript in the `TRAK2` 
gene. The transcript names are stored in the `tx_name` metadata column. We will 
need them to subset the extracted intron and exon regions:

```{r}
(trak2_tx_names <- mcols(trak2_txs)$tx_name)
```

The exon and intron genomic ranges for all the transcripts in the gene model 
can be extracted with the `exonsBy` and `intronsByTranscript` functions, 
respectively. Both functions return a `GRangesList` object. Then we keep only 
the exon and intron for the transcripts of the `TRAK2` gene by subsetting 
each `GRangesList` object by the `TRAK2` transcript names.

Extract the exon regions:

```{r}
trak2_exbytx <- exonsBy(txdb, "tx", use.names = TRUE)[trak2_tx_names]
elementNROWS(trak2_exbytx)
```

... and the intron regions:

```{r}
trak2_inbytx <- intronsByTranscript(txdb, use.names = TRUE)[trak2_tx_names]
elementNROWS(trak2_inbytx)
```

Next we want the DNA sequences for these exons and introns. The `getSeq` 
function from the `r Biocpkg("Biostrings")` package can be used to query a 
`BSgenome` object with a set of genomic ranges and retrieve the corresponding 
DNA sequences.

```{r}
suppressPackageStartupMessages({
    library(BSgenome.Hsapiens.UCSC.hg38)
})
```

Extract the exon sequences:

```{r}
(trak2_ex_seqs <- getSeq(Hsapiens, trak2_exbytx))
```

... and the intron sequences:

```{r}
(trak2_in_seqs <- getSeq(Hsapiens, trak2_inbytx))
```

Each list element is a `DNAStringSet` object, e.g.:

```{r}
trak2_in_seqs[[1]]
```

Each sequence here corresponds to a single intron.

# Further reading

- The `r Biocpkg("GenomicFeatures")` vignette.

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>
