---
title: How to read gene sets from GMT files
author: Robert Castelo
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to read gene sets from GMT files}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

A **gene set** is simple yet useful way to define pathways without regard
to the specific molecular interactions. It can also represent the signature
of a characteristic pattern of gene expression associated with a cell type
or a disease for diagnostic or prognostic purposes. An important source of
gene sets is the
[Molecular Signatures Database (MSigDB)](https://www.gsea-msigdb.org/gsea/msigdb),
which stores them as plain text files following the so-called
[_gene matrix transposed_ (GMT) format](https://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29).
In the GMT format, each line stores a gene set with the following values
separated by tabs:

  * A unique gene set identifier.
  * A gene set description.
  * One or more gene identifiers.

Because each different gene set may consist of a different number of genes, each
line in a GMT file may contain a different number of tab-separated values. This
means that the GMT format is not a tabular format, and therefore cannot be directly
read with base R functions such as `read.table()` or `read.csv()`.

# Bioconductor packages used in this document

* `r Biocpkg("GSEABase")`
* `r Biocpkg("GSVA")`

# How to read gene sets from GMT files

The package `r Biocpkg("GSEABase")` provides object classes and methods to represent
and manipulate gene sets, including reading GMT files with the function `getGmt()`.
Here is a first example in which we read a GMT file from the MSigDB database:

```{r, message=FALSE}
library(GSEABase)

URL <- paste("https://data.broadinstitute.org/gsea-msigdb/msigdb/release",
             "2023.2.Hs/c7.immunesigdb.v2023.2.Hs.symbols.gmt", sep="/")
genesets <- getGmt(URL, geneIdType=SymbolIdentifier())
genesets
length(genesets)
```
Next to the filename of the GMT file, we provided the argument
`geneIdType=SymbolIdentifier()`. This argument adds the necessary metadata that
later on allows other software to figure out what kind of gene identifiers are
used in this collection of gene sets, to attempt mapping them to other type of
identifiers, if necessary. While this argument is optional, we should always
try to provide it.

The output of the `getGmt()` function is a `GeneSetCollection` object, which
pretty much behaves like a list object and many Bioconductor packages analysing
gene sets may expect that kind of object as input. However, we can coerce it to
a regular `list` object with the function `geneIds()`:

```{r}
genesets.list <- geneIds(genesets)
head(lapply(genesets.list, head, 3), 3)
```
By definition, GMT files should have unique gene set identifiers and if we
attempt to read with `getGmt()` a malformed GMT file with duplicated
identifiers it will prompt an error:

```{r, error=TRUE}
URL <- paste("https://data.broadinstitute.org/gsea-msigdb/msigdb/release",
             "7.5/c2.all.v7.5.symbols.gmt", sep="/")
genesets <- getGmt(URL, geneIdType=SymbolIdentifier())
```
So if we still want to read the contents of that GMT file, we would have to
download it and edit the lines with duplicated identifiers. An alternative is
to use the function `readGMT()` of the `r Biocpkg("GSVA")` package:

```{r, message=FALSE}
library(GSVA)

genesets <- readGMT(URL)
genesets
```
As we may see, it has detected the presence duplicated gene set identifiers and
applied a deduplication policy, which can be changed through the parameter
`dedupUse`; see its help page with `?readGMT` for further deduplication
policies. As `getGmt()`, it also takes a parameter `geneIdType`, but by default
it will automatically detect the type of gene identifier and fill in the
corresponding metadata in the resulting `GeneSetCollection` object, so that in
principle we need not to specify it.

# Further reading

- The [Introduction to GSEABase](https://bioconductor.org/packages/release/bioc/vignettes/GSEABase/inst/doc/GSEABase.html)
  vignette from the `r Biocpkg("GSEABase")` package.
- The help pages of the `getGmt()` and `readGMT()` functions in the
  `r Biocpkg("GSEABase")` and `r Biocpkg("GSVA")` packages, respectively.
- The Bioconductor package `r Biocpkg("msigdb")` provides an API for
  directly querying and downloading MSigDB GMT files as `GeneSetCollection`
  objects.
- Similarly to `r Biocpkg("msigdb")`, the `r CRANpkg("msigdbr")` CRAN
  package provides an API to query and download MSigDB GMT files, but as
  [tidy](https://r4ds.hadley.nz/data-tidy.html) `data.frame` objects
  with one gene per row.

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>
