---
title: How to compute read coverage
author: Bioconductor Core Team
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to compute read coverage}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

This HowTo has been adapted from the list of HowTos provided in the 
[vignette](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesHOWTOs.pdf)
for the `r Biocpkg("GenomicRanges")` Bioconductor package. 

# Bioconductor packages used in this document

* `r Biocpkg("pasillaBamSubset")`
* `r Biocpkg("GenomicAlignments")`

# How to compute read coverage

By "read coverage", we mean the number of reads that cover a given genomic position. 
Computing the read coverage generally consists in computing the coverage at 
each position in the genome.
This can be done with the `coverage` function from `r Biocpkg("GenomicAlignments")`.
For illustration, we use data from the `r Biocpkg("pasillaBamSubset")` data 
package.

```{r}
suppressPackageStartupMessages({
    library(pasillaBamSubset)
})
# Path to a bam file with single-end reads
(un1 <- untreated1_chr4())
```

Single-end reads can be read with the `readGAlignments` function from 
`r Biocpkg("GenomicAlignments")`.

```{r}
suppressPackageStartupMessages({
    library(GenomicAlignments)
})
(reads <- readGAlignments(un1))
```

The coverage can then be calculated using the `coverage` function from 
`r Biocpkg("GenomicAlignments")`: 

```{r}
(cvg <- coverage(reads))
```

We can extract the coverage on a specific chromosome: 

```{r}
cvg$chr4
```

We can also perform calculations on the returned object, e.g. to get the 
average and maximum coverage on chromosome 4:

```{r}
mean(cvg$chr4)
max(cvg$chr4)
```

We can use the `slice` function to find the genomic regions where the coverage
is greater or equal to a given threshold.

```{r}
(chr4_highcov <- slice(cvg$chr4, lower = 500))
```

The *weight* of a given such region can be defined as the number of aligned 
nucleotides that belong to the region (i.e. the area under the coverage curve). 
It can be obtained with `sum`:

```{r}
sum(chr4_highcov)
```

Note that `coverage` and `slice` are generic functions with methods for 
different types of objects. See `?coverage` or `?slice` for more information.


# Further reading

- The `r Biocpkg("GenomicAlignments")` vignette.

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>
