---
title: How to compute sequence composition for genomic regions
author: EDIT:HowToAuthor
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to compute sequence composition for genomic regions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
format:
    html:
        toc: true
        html-math-method: mathjax
        embed-resources: true
---

```{r}
#| echo: false

suppressPackageStartupMessages({
    library(BiocStyle)
})
```

This HowTo shows how to fetch sequences for given regions from a reference and
compute composition features for them.

# Bioconductor packages used in this document

* `r Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")`
* `r Biocpkg("BSgenome.Hsapiens.UCSC.hg38")`
* `r Biocpkg("GenomicFeatures")`
* `r Biocpkg("Biostrings")`

# How to compute sequence composition for genomic regions

Mammalian genomes contain regulatory regions with a high content of CpG 
dinucleotides, so called *CpG islands*, a side effect indirectly caused by
the methylation of cytosines in CpG contexts. Many transcripts contain
a CpG island in their promoter region, near the start of the transcript.
In this HowTo, we will identify such transcripts by analyzing the CpG
composition of promoter sequences.

We first obtain the coordinates of promoter regions from the
`r Biocpkg("TxDb.Hsapiens.UCSC.hg38.knownGene")` package, using the
`promoters` function from the `r Biocpkg("GenomicFeatures")` package:

```{r}
suppressPackageStartupMessages({
    library(GenomicFeatures)
    library(TxDb.Hsapiens.UCSC.hg38.knownGene)
})
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
prom <- promoters(txdb, upstream = 1000, downstream = 1000)
prom
```

Here, promoter regions are defined as 2000 base pair regions centered on
transcript start sites (the 5'-end of a transcript). The warning message above
is caused by start sites that are near the boundary of a chromosome, such that
the promoter region will extend beyond its beginning or end. By calling `trim`
on the `GRanges` object, all regions are truncated to the limits of the
chromosomes, resulting in a few promoters shorter than 2000 base pairs:

```{r}
summary(width(prom) == 2000)
prom <- trim(prom)
summary(width(prom) == 2000)
```

It may be useful to now subset the extracted promoters, for example to only
retain promoters on autosomes, or only a single promoter per gene.
Here, we will remove redundant promoters resulting from transcripts with
identical start sites:

```{r}
summary(duplicated(prom))
prom <- prom[!duplicated(prom)]
```

Next, we extract the sequences of the promoters from the reference genome
assembly in the `r Biocpkg("BSgenome.Hsapiens.UCSC.hg38")` package:

```{r}
suppressPackageStartupMessages({
    library(Biostrings)
    library(BSgenome.Hsapiens.UCSC.hg38)
})
promseq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, prom)
promseq
```

Alternative sources of the genome sequence are also supported by the
`getSeq` function, such as `DNAStringSet` objects or files in fasta or 2bit
format:

```{r}
showMethods(getSeq)
```

Finally, we calculate some sequence features: The percent of G and C bases, and
the ratio of observed over expected CpG dinucleotides (where the expected
frequency of a dinucleotide is the product of the frequencies of the two
mononucleoties). We obtain the mono- and dinucleotide frequencies from the
sequences using the `oligonucleotideFrequency` function from the
`r Biocpkg("Biostrings")` package:

```{r}
mo <- oligonucleotideFrequency(promseq, width = 1, as.prob = TRUE)
di <- oligonucleotideFrequency(promseq, width = 2, as.prob = TRUE)

head(mo)
head(di)

percentGC <- 100 * (mo[, "C"] + mo[, "G"])
expectedCpG <- mo[, "C"] * mo[, "G"]
obsExpRatioCpG <- di[, "CG"] / expectedCpG
```

Finally, we can look at the distribution of these sequence features.
The percentage of G+C bases in promoters spans a broad range from about 30% to
over 70%:

```{r}
#| fig.width: 6
#| fig.height: 6
hist(percentGC, 60)
```

Defining CpG island promoters using this distribution would be hard, but is
much easier using the observed over expected ratio of CpGs, which shows a
much clearer bimodal structure:

```{r}
#| fig.width: 6
#| fig.height: 6
hist(obsExpRatioCpG, 60)
abline(v = 0.45, col = 2, lty = 2)
```

Interestingly, almost all ratios are below 1.0, indicating that CpG islands
are in fact not enriched for CpG dinucleotides, but rather lose them at a
lower rate than other genomic regions.

Plotting the ratio versus the GC percentage furthermore reveals that CpG island
promoters are also slightly GC-richer:

```{r}
#| fig.width: 6
#| fig.height: 6
plot(obsExpRatioCpG, percentGC, pch = "*", col = "#22222205")
abline(v = 0.45, col = 2, lty = 2)
```

# Further reading

More examples of how to represent and work with biological sequences are
contained in the vignettes of the `r Biocpkg("Biostrings")` package,
available for example from `vignette(package = "Biostrings")`.

For more information about CpG islands, see:

- Bird A, Taggart M, Frommer M, Miller OJ, Macleod D. A fraction of the mouse
  genome that is derived from islands of nonmethylated, CpG-rich DNA.
  Cell. 1985 Jan;40(1):91-9. doi: 10.1016/0092-8674(85)90312-5. PMID: 2981636.
- Gardiner-Garden M, Frommer M. CpG islands in vertebrate genomes.
  J Mol Biol. 1987 Jul 20;196(2):261-82.
  doi: 10.1016/0022-2836(87)90689-9. PMID: 3656447.

# Session info

<details>
<summary><b>
Click to display session info
</b></summary>
```{r}
sessionInfo()
```
</details>
