---
title: "How to extend S4Vectors::Vector with a small custom class"
author:
- "BiocHowTo contributors"
date: "2026-03-12"
package: "BiocHowTo"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{How to extend S4Vectors::Vector with a small custom class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Goal

This How-To shows a minimal, practical pattern for extending a *core*
Bioconductor S4 **VIRTUAL** class: `?S4Vectors::Vector-class`.

We implement a `Percentage` class that:

- stores values as **decimals** between zero and one (inclusive)
- prints as a **percent** with a configurable number of decimal places
- supports a small set of *Vector-like* behaviors (length, show, ordering,
  coercions)
- demonstrates the typical "extension checklist" for Bioconductor S4 classes

# Prerequisites

```{r}
library(S4Vectors)
library(scales)
```

Note. We use `scales` only for pretty printing.

# A checklist for extending `Vector`

When extending the `Vector` virtual class in `S4Vectors`, a developer should
check the following list of steps to ensure a well-behaved class:

1. Class definition
2. Validity
3. User-facing constructor
4. Parallel slot names
5. show() method
6. Setters / Getters (for new slots)
7. Coercions
8. Optional: Ordering (with `order`)
9. Optional: comparison helpers (e.g., `sameAsPreviousROW`)

The `Percentage` class below follows these steps.

# Step 1: Define the class

We define an S4 class that *contains* the `Vector` virtual class and adds two
slots:

- `decimal`: the numeric decimals
- `nplaces`: how many decimal places to show when printing

```{r}
.Percentage <- setClass(
    "Percentage",
    contains = "Vector",
    slots = list(
        decimals = "numeric",
        nplaces = "integer"
    )
)
```

# Step 2: Add validity checks

We ensure:

- decimal **values** are between 0 and 1
- `nplaces` is a scalar integer of the number of decimal **places**

```{r}
.validate_Percentage <- function(x) {
    msg <- NULL
    if (any(x@decimals > 1.0 | x@decimals < 0.0, na.rm = TRUE))
        msg <- c(msg, "All decimal values must be between 0 and 1")
    if (length(x@nplaces) != 1L || !is.integer(x@nplaces) || is.na(x@nplaces))
        msg <- c(msg, "The number of decimal places must be a single integer")
    msg
}

setValidity2("Percentage", .validate_Percentage)
```

# Step 3: Create a user-facing constructor

A constructor function provides a user-friendly interface to create objects of
the class. It should call the internal class constructor and that, in turn,
will ensure that the object is valid.

```{r}
Percentage <- function(values, nplaces = 1L) {
    .Percentage(
        decimals = values, nplaces = as.integer(nplaces)
    )
}
```

# Step 4: Declare parallel slots

Many Bioconductor vector-like classes support the concept of "parallel slots"
that should be subset together or combined.

```{r}
setMethod(
    "parallel_slot_names", "Percentage",
    function(x) c("decimals", callNextMethod())
)
```

# Step 5: Define printing (`show`)

```{r}
.show_Percentage <- function(object) {
    cat("Percentage-object (length =", length(object), "):\n")
    cat(scales::percent(object@decimals, accuracy = 1 / 10^object@decimals))
}

setMethod("show", "Percentage", .show_Percentage)
```

# Step 6: Add basic accessors (getters / setters)

## Access fractions

In the issue example, `score()` is used to extract `vec`. We keep that here.

```{r}
setMethod("score", "Percentage", function(x) x@decimals)
```

## Get/set decimals

```{r}
setGeneric("places", function(x, ...) standardGeneric("places"))
setMethod("places", "Percentage", function(x) x@nplaces)

setGeneric("places<-", function(x, value) standardGeneric("places<-"))
setReplaceMethod("places", "Percentage", function(x, value) {
    x@nplaces <- as.integer(value)
    validObject(x)
    x
})
```

# Step 7: Coercion methods

Provide a few convenient conversions.

```{r}
setAs("Percentage", "integer", function(from) as.integer(from@decimals * 100))
setAs("Percentage", "numeric", function(from) from@decimals)
setAs("Percentage", "character", function(from) {
    scales::percent(from@decimals, accuracy = 1 / 10^from@decimals)
})
setAs("numeric", "Percentage", function(from) Percentage(from))
```

# Step 8 (optional): Support ordering (and “same as previous row”)

If you want your class to participate naturally in sorting workflows, define `order()`.

```{r}
.orderPercentage <- function(
    ...,
    na.last = TRUE,
    decreasing = FALSE,
    method = c("auto", "shell", "radix")
) {
    args <- list(...)
    if (length(args) != 1L)
        stop("Ordering multiple Percentage-objects not currently supported")
    order(
        args[[1]]@decimals,
        na.last = na.last,
        decreasing = decreasing,
        method = method
    )
}

setMethod("order", "Percentage", .orderPercentage)
```

Some Bioconductor internals and packages use `sameAsPreviousRow()` for grouping/comparison in vector-like contexts. If you need it, implement it by delegating to the underlying numeric vector.

**Note:** the function name in the issue appears as `sameAsPreviousROW` (uppercase `ROW`). In Bioconductor code this is typically `sameAsPreviousRow` (lowercase `ow`). If you decide to add this, confirm the correct generic name for your context.

```{r, eval=FALSE}
setMethod(
    "sameAsPreviousRow", "Percentage", function(x) sameAsPreviousRow(x@decimals)
)
```

# Try it

```{r}
p <- Percentage(c(0.1, 0.25, 0.5), nplaces = 1L)
p
score(p)

places(p) <- 3
p

as(p, "character")
as(p, "integer")
order(p)
```

# Notes and common pitfalls

- **Validity should be strict**: it saves debugging time later.
- **Parallel slots**: if you add additional data slots that must be subsetted
  alongside the core data, include them in `parallel_slot_names()`.
- **Decide what “Vector-like” means for your class**: for some extensions you
  may also need methods for subsetting, combining, `length`, `[` / `[[`, `c`,
  etc. (many are inherited, but not always sufficient depending on slot
  structure).
- **Naming**: consider using a public class name that won't collide with
  existing packages.

# Session info

```{r}
sessionInfo()
```
